---
title: "Analysis"
author: "Hayden Simm"
date: "2026-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(car)
library(effects)
```

Loading in dataframes 

```{r}
killi_1<-read_csv("Data_Set/HS_MD_killifish.csv")
killi 
killi<- killi %>%
  drop_na(where(is.numeric))

str(killi)
```


Changing predictor Acclimation temperature and Acclimation salinity to factors 
```{r}
killi$Accliamtion_Temperature<- factor (killi$Accliamtion_Temperature)
killi$Acclimation_Salinity<- factor(killi$Acclimation_Salinity)
killi$Assay_Temperature<- factor(killi$Assay_Temperature)

str(killi)
```
Make the enzymes to long form, so we can plot all three at once
```{r}
killi_long <- killi %>%
  pivot_longer(
    cols = c(COX_Activity, LDH_Activity, CS_Activity),
    names_to = "Enzyme",  
    values_to = "Activity" 
  )
str(killi_long)
```


Plotting assay temp and enzymes grouped by acclimation temperature
```{r}
ggplot(killi_long, aes(x = Assay_Temperature, y = log(Activity), fill = Accliamtion_Temperature)) +
geom_boxplot() +
  facet_wrap(vars(Enzyme), ncol=1) +
  theme_minimal() +
  xlab("Assay Temperature") +
  ylab("Enzyme Activity") +
  labs(fill = "Acclimation Temp")

```

Plotting assay temperature enzyme activities grouped by acclimation temperature 

```{r}
cox_activity<- ggplot(killi, aes(x = Assay_Temperature, y = COX_Activity, fill =Accliamtion_Temperature)) +
geom_boxplot()  +
  theme_minimal() +
  xlab("Assay Temperature") +
  ylab("COX Activity") +
  labs(fill = "Acclimation Temp")

cs_activity<- ggplot(killi, aes(x = Assay_Temperature, y = CS_Activity, fill = Accliamtion_Temperature)) +geom_boxplot()  +
  theme_minimal() +
  xlab("Assay Temperature") +
  ylab("CS Activity") +
  labs(fill = "Acclimation Temp")

LDH_activity<- ggplot(killi, aes(x = Assay_Temperature, y = LDH_Activity, fill = Accliamtion_Temperature)) +geom_boxplot()  +
  theme_minimal() +
  xlab("Assay Temperature") +
  ylab("LDH Activity") +
  labs(fill = "Acclimation Temp")



library(patchwork)
cox_activity+cs_activity

LDH_activity
```

Plotting assay temperature by body condition
```{r}
ggplot(killi_long, aes(x= Assay_Temperature, y= Body_Condition, fill= Accliamtion_Temperature)) +
  geom_boxplot() + theme_minimal() +
  xlab("Assay Temperature") +
  ylab("Body Condition") +
  labs(fill = "Acclimation Temp")
```


Grouping body condition by salinity acclimation 
```{r}
##grouping by acclimation salinity 

ggplot(killi_long, aes(x= Assay_Temperature, y= Body_Condition, fill= Acclimation_Salinity)) +
  geom_boxplot() + theme_minimal() +
  xlab("Assay Temperature") +
  ylab("Body Condition") +
  labs(fill = "Acclimation Temp")
```

```
multivariate response equation for the different enzymes and the impact acclimation temperature and salinity, assay temp, each fish has on the enzyme expression. Tried to add Fish_ID into the MANOVA but could not because it could not do the repeated measures with fish. 

```{r}

Manova_enzymes<- manova(cbind(COX_Activity,CS_Activity,LDH_Activity)~ (Accliamtion_Temperature + Acclimation_Salinity)*Assay_Temperature, data=killi)


Anova(Manova_enzymes)

summary(Manova_enzymes)


```

Linear model for body condition with predictors temperature acclimation, salinity acclimation, and temperature assay. Wanted Fish ID but could not have it in the plot...
```{r}
library(lme4)

body<- 
  lm(Body_Condition ~ (Accliamtion_Temperature + Acclimation_Salinity)*Assay_Temperature, data=killi)

summary(body)

plot(body)


body2<- 
  lm(Body_Condition ~ Accliamtion_Temperature* Acclimation_Salinity* Assay_Temperature, data=killi)
  
plot(body2)

```
`

all effects plot 
```{r}

plot(allEffects(body))

plot(allEffects(body2))
```

Hematcrit 
```{r}

hem<- 
  lm(Hematocrit~ (Accliamtion_Temperature + Acclimation_Salinity)*Assay_Temperature, data=killi)



```

