---
title: "Analysis"
author: "Hayden Simm"
date: "2026-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(car)
```

Loading in dataframes 

```{r}
killi<-read_csv("Data_Set/HS_MD_killifish.csv")

str(killi)

```

Changing predictor Acclimation temperature and Acclimation salinity to factors 
```{r}
killi$Accliamtion_Temperature<- factor (killi$Accliamtion_Temperature)
killi$Acclimation_Salinity<- factor(killi$Acclimation_Salinity)
killi$Assay_Temperature<- factor(killi$Assay_Temperature)
str(killi)
```
Make the enzymes to long form, so we can plot all three at once
```{r}
killi_long <- killi %>%
  pivot_longer(
    cols = c(COX_Activity, LDH_Activity, CS_Activity),
    names_to = "Enzyme",  
    values_to = "Activity" 
  )
head(killi_long)
```


Plotting assay temp and enzymes grouped by acclimation temperature
```{r}
ggplot(killi_long, aes(x = Assay_Temperature, y = log(Activity), fill = Accliamtion_Temperature)) +
geom_boxplot() +
  facet_wrap(vars(Enzyme), ncol=1) +
  theme_minimal() +
  xlab("Assay Temperature") +
  ylab("Enzyme Activity") +
  labs(fill = "Acclimation Temp")

```
Plotting assay temperature enzyme activities grouped by acclimation temperature 

```{r}
cox_activity<- ggplot(killi, aes(x = Assay_Temperature, y = COX_Activity, fill =Accliamtion_Temperature)) +
geom_boxplot()  +
  theme_minimal() +
  xlab("Assay Temperature") +
  ylab("COX Activity") +
  labs(fill = "Acclimation Temp")

cs_activity<- ggplot(killi, aes(x = Assay_Temperature, y = CS_Activity, fill = Accliamtion_Temperature)) +geom_boxplot()  +
  theme_minimal() +
  xlab("Assay Temperature") +
  ylab("CS Activity") +
  labs(fill = "Acclimation Temp")

LDH_activity<- ggplot(killi, aes(x = Assay_Temperature, y = LDH_Activity, fill = Accliamtion_Temperature)) +geom_boxplot()  +
  theme_minimal() +
  xlab("Assay Temperature") +
  ylab("LDH Activity") +
  labs(fill = "Acclimation Temp")

library(patchwork)
cox_activity+cs_activity

LDH_activity
```

Plotting assay temperature by body condition
```{r}
ggplot(killi_long, aes(x= Assay_Temperature, y= Body_Condition, fill= Accliamtion_Temperature)) +
  geom_boxplot() + theme_minimal() +
  xlab("Assay Temperature") +
  ylab("Body Condition") +
  labs(fill = "Acclimation Temp")
```
`


```


```
multivariate response equation for the different enzymes and the impact acclimation temperature and salinity, assay temp, each fish has on the enzyme expression. Have fish ID as fixed effect. Think we need type ll manova from CAR package
```{r}

Manova_enzymes<- manova(cbind(COX_Activity,CS_Activity,LDH_Activity)~ (Accliamtion_Temperature + Acclimation_Salinity + Fish_ID)*Assay_Temperature, data=killi)

Anova(Manova_enzymes)

summary(Manova_enzymes)




```

Linear model for body condition with predictors temperature acclimation, salinity acclimation, and temperature assay
```{r}
library(lme4)
body<- lmer(Body_Condition ~ (Accliamtion_Temperature + Acclimation_Salinity)*Assay_Temperature +(1+Assay_Temperature|Fish_ID),  data=killi)

summary(body)
 

plot(body)

```


```{r}


library(effects)

plot(allEffects(body))

```

