---
title: "Analysis"
author: "Hayden Simm"
date: "2026-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(car)
library(heplots)
library(jmv)
```

Loading in dataframes 

```{r}
killi<-read_csv("../Data_Set/HS_MD_killifish.csv")
str(killi)

```

Changing predictor Acclimation temperature and Acclimation salinity to factors 
```{r}
killi$Acclimation_Temperature<- factor (killi$Acclimation_Temperature)
killi$Acclimation_Salinity<- factor(killi$Acclimation_Salinity)
killi$Assay_Temperature<- factor(killi$Assay_Temperature)
str(killi)
```

```
Plotting assay temperature enzyme activities grouped by acclimation temperature 

GGplots for COX, have the points overlaid on the boxplots
```{r}

cox_activity <- ggplot(
  killi,
  aes(
    x = Acclimation_Temperature,
    y = COX_Activity,
    fill = Assay_Temperature
  )
) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_jitter(
    position = position_jitterdodge(
      dodge.width = 0.8,
      jitter.width = 0.2
    ),
    alpha = 0.7,
    size = 2,
    shape = 21  
) +
  scale_fill_manual(
    values = c(
      "10" = "#6BAED6",
      "24" = "#FB6A4A"
    )
  ) +
  theme_classic()+
  labs(
    x = "Acclimation Temperature",
    y = "COX Activity",
    fill="Assay Temperature"
  ) +
   theme(
     panel.grid = element_blank(),                        
    axis.title = element_text(size = 12), # axis labels bigger & bold
    strip.text = element_text(size = 12), # facet labels bigger & bold
    legend.title = element_text(size = 12))#
  
plot(cox_activity)
```
Saving photo
```{r}

ggsave(
  filename = "cox_activity.png",
  plot = cox_activity,
  width = 8,    # in inches
  height = 5,
  dpi = 300
)
```

LDH
```{r}
LDH_activity <- ggplot(
  killi,
  aes(
    x = Acclimation_Temperature,
    y = LDH_Activity,
    fill = Assay_Temperature
  )
) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_jitter(
    position = position_jitterdodge(
      dodge.width = 0.8,
      jitter.width = 0.2
    ),
    alpha = 0.7,
    size = 2,
    shape = 21  
) +
  scale_fill_manual(
    values = c(
      "10" = "#6BAED6",
      "24" = "#FB6A4A"
    )
  ) +
  theme_classic()+
  labs(
    x = "Acclimation Temperature",
    y = "LDH Activity",
    fill="Assay Temperature"
  ) +
   theme(
     panel.grid = element_blank(),                        
    axis.title = element_text(size = 12), # axis labels bigger & bold
    strip.text = element_text(size = 12), # facet labels bigger & bold
    legend.title = element_text(size = 12))#
  
plot(LDH_activity)
```
Saving photos 
```{r}
ggsave(
  filename = "LDH_activity.png",
  plot = LDH_activity,
  width = 8,    
  height = 5,
  dpi = 300
)
```

CS
```{r}
CS_activity <- ggplot(
  killi,
  aes(
    x = Acclimation_Temperature,
    y = CS_Activity,
    fill = Assay_Temperature
  )
) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_jitter(
    position = position_jitterdodge(
      dodge.width = 0.8,
      jitter.width = 0.2
    ),
    alpha = 0.7,
    size = 2,
    shape = 21  
) +
  scale_fill_manual(
    values = c(
      "10" = "#6BAED6",
      "24" = "#FB6A4A"
    )
  ) +
  theme_classic()+
  labs(
    x = "Acclimation Temperature",
    y = "CS Activity",
    fill="Assay Temperature"
  ) +
   theme(
     panel.grid = element_blank(),                        
    axis.title = element_text(size = 12), # axis labels bigger & bold
    strip.text = element_text(size = 12), # facet labels bigger & bold
    legend.title = element_text(size = 12))#
  
plot(CS_activity)
```

Saving photo
```{r}
ggsave(
  filename = "CS_activity.png",
  plot = CS_activity,
  width = 8,  
  height = 5,
  dpi = 300
)
```



```{r}
Body_Condition <- ggplot(
  killi,
  aes(
    x = Acclimation_Temperature,
    y = Body_Condition,
    fill = Assay_Temperature
  )
) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_jitter(
    position = position_jitterdodge(
      dodge.width = 0.8,
      jitter.width = 0.2
    ),
    alpha = 0.7,
    size = 2,
    shape = 21  
) +
  scale_fill_manual(
    values = c(
      "10" = "#6BAED6",
      "24" = "#FB6A4A"
    )
  ) +
  theme_classic()+
  labs(
    x = "Acclimation Temperature",
    y = "Body Condition",
    fill="Assay Temperature"
  ) +
   theme(
     panel.grid = element_blank(),                        
    axis.title = element_text(size = 12), # axis labels bigger & bold
    strip.text = element_text(size = 12), # facet labels bigger & bold
    legend.title = element_text(size = 12))#
  
plot(Body_Condition)

```
Saving photo
```{r}
ggsave(
  filename = "Body_Condition.png",
  plot = Body_Condition,
  width = 8,  
  height = 5,
  dpi = 300
)
```


Logging the enzyme activites 
```{r}
killi_log<-killi  %>% mutate(across(c(CS_Activity, COX_Activity, LDH_Activity),
                ~log(.x),
                .names = "log_{.col}"))

str(killi_log)
```


MANCOVA- because body condition would be a continuous predictor. 

Checking the assumption of no mulitcolinearity 
```{r}
cor(killi_log[, c("log_CS_Activity","log_COX_Activity","log_LDH_Activity")])
pairs(killi_log[, c("log_CS_Activity","log_COX_Activity","log_LDH_Activity")])
```
Checking assunption of regression slopes
```{r}
Group <- interaction(killi_log$Acclimation_Temperature,
                        killi_log$Acclimation_Salinity,
                        killi_log$Assay_Temperature)
                        
CS <- lm(log_CS_Activity ~ Group * Body_Condition, data = killi_log)
anova(CS)

COX <- lm(log_COX_Activity ~ Group * Body_Condition, data = killi_log)
anova(COX)

LDH <- lm(log_LDH_Activity ~ Group * Body_Condition, data = killi_log)
anova(LDH)
```
All assumptions are met

Making the MANCOVA 
```{r}
jmv::mancova(data= killi_log, deps= c(log_COX_Activity,log_CS_Activity, log_LDH_Activity), factors= c(Acclimation_Temperature, Acclimation_Salinity, Assay_Temperature), multivar= list('pillai'),  covs = Body_Condition, boxM = TRUE,
  shapiro = TRUE, qqPlot = TRUE)
```
The qq-plot demonstrates the data follows normal distrubution even though shapiro wilks test says the data does not






EXTRA FROM BEOFRE 
Predictors: Acclimation temperautre, acclimation salinity, body condition. Body condtion is a continous predictor while the others are all factors.

We want to see if there the interaction among these three predictors has an impact on the enzyme activity.
```{r}

Manova_enzymes<- manova(cbind(log_COX_Activity,log_CS_Activity,log_LDH_Activity)~ Acclimation_Temperature * Acclimation_Salinity * Assay_Temperature * Body_Condition, data=killi_log)

Anova(Manova_enzymes)

summary(Manova_enzymes)

heplot(Manova_enzymes)

```

Checking assumptions of MANOVA
```{r}

                        
boxM(cbind(log_COX_Activity,log_CS_Activity,log_LDH_Activity) ~ Group, data = killi_log) 



scatterplotMatrix(~ log_COX_Activity + log_CS_Activity + log_LDH_Activity,
                  data = killi_log)
```


```