---
title: "Analysis"
author: "Hayden Simm"
date: "2026-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(car)
```

Loading in dataframes 

```{r}
killi<-read_csv("../Data_Set/HS_MD_killifish.csv")
str(killi)

```

Changing predictor Acclimation temperature and Acclimation salinity to factors 
```{r}
killi$Acclimation_Temperature<- factor (killi$Acclimation_Temperature)
killi$Acclimation_Salinity<- factor(killi$Acclimation_Salinity)
killi$Assay_Temperature<- factor(killi$Assay_Temperature)
str(killi)
```

```
Plotting assay temperature enzyme activities grouped by acclimation temperature 

GGplots for COX, have the points overlaid on the boxplots
```{r}

cox_activity <- ggplot(
  killi,
  aes(
    x = Acclimation_Temperature,
    y = COX_Activity,
    fill = Assay_Temperature
  )
) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_jitter(
    position = position_jitterdodge(
      dodge.width = 0.8,
      jitter.width = 0.2
    ),
    alpha = 0.7,
    size = 2,
    shape = 21  
) +
  scale_fill_manual(
    values = c(
      "10" = "#6BAED6",
      "24" = "#FB6A4A"
    )
  ) +
  theme_classic()+
  labs(
    x = "Acclimation Temperature",
    y = "COX Activity",
    fill="Assay Temperature"
  ) +
   theme(
     panel.grid = element_blank(),                        
    axis.title = element_text(size = 12), # axis labels bigger & bold
    strip.text = element_text(size = 12), # facet labels bigger & bold
    legend.title = element_text(size = 12))#
  
plot(cox_activity)
```
Saving photo
```{r}

ggsave(
  filename = "cox_activity.png",
  plot = cox_activity,
  width = 8,    # in inches
  height = 5,
  dpi = 300
)
```

LDH
```{r}
LDH_activity <- ggplot(
  killi,
  aes(
    x = Acclmation_Temperature,
    y = LDH_Activity,
    fill = Assay_Temperature
  )
) +
  geom_boxplot(position = position_dodge(width = 0.8)) + 
  geom_jitter(
    position = position_jitterdodge(
      dodge.width = 0.8,
      jitter.width = 0.2
    ),
    alpha = 0.7,
    size = 2,
    shape = 21  
)
  facet_wrap(
  ~ Acclimation_Salinity,
  nrow = 1,
  labeller = labeller(
    Acclimation_Salinity = c(
      "11" = "Low Salinity",
      "28" = "High Salinity"
    )
  )
) +
  scale_fill_manual(
    values = c(
      "10" = "#6BAED6",
      "24" = "#FB6A4A"
    )
  ) +
  theme_classic()+
  labs(
    x = "Acclimation Temperature",
    y = "LDH Activity",
    fill="Assay Temperature"
  ) +
   theme(
     panel.grid = element_blank(),                        
    axis.title = element_text(size = 12), 
    strip.text = element_text(size = 12), 
    legend.title = element_text(size = 12))
  
plot(LDH_activity)
```
Saving photos 
```{r}
ggsave(
  filename = "LDH_activity.png",
  plot = LDH_activity,
  width = 8,    
  height = 5,
  dpi = 300
)
```

CS
```{r}
CS_activity <- ggplot(
  killi,
  aes(
    x = Acclimation_Temperature,
    y = CS_Activity,
    fill = Assay_Temperature
  )
) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(
  ~ Acclimation_Salinity,
  nrow = 1,
  labeller = labeller(
    Acclimation_Salinity = c(
      "11" = "Low Salinity",
      "28" = "High Salinity"
    )
  )
) +
  scale_fill_manual(
    values = c(
      "10" = "#6BAED6",
      "24" = "#FB6A4A"
    )
  ) +
  theme_classic()+
  labs(
    x = "Acclimation Temperature",
    y = "CS Activity",
    fill="Assay Temperature"
  ) +
   theme(
     panel.grid = element_blank(),                        
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12), 
    legend.title = element_text(size = 12))
  
plot(CS_activity)
```

Saving photo
```{r}
ggsave(
  filename = "CS_activity.png",
  plot = CS_activity,
  width = 8,  
  height = 5,
  dpi = 300
)
```



```{r}
Body_Condition<- ggplot(
  killi,
  aes(
    x = Acclimation_Temperature,
    y = Body_Condition
  )
) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(
  ~ Acclimation_Salinity,
  nrow = 1,
  labeller = labeller(
    Acclimation_Salinity = c(
      "11" = "Low Salinity",
      "28" = "High Salinity"
    )
  )) +
  theme_classic()+
  labs(
    x = "Acclimation Temperature",
    y = "Body Condition",
  ) +
   theme(
     panel.grid = element_blank(),                    
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12), 
    legend.title = element_text(size = 12))
  
plot(Body_Condition)

```
Saving photo
```{r}
ggsave(
  filename = "Body_Condition.png",
  plot = Body_Condition,
  width = 8,  
  height = 5,
  dpi = 300
)
```
multivariate response equation for the different enzymes and the impact acclimation temperature and salinity, assay temp, each fish has on the enzyme expression. Tried to add Fish_ID into the MANOVA but could not because it could not do the repeated measures with fish. 

Manova set up: 

Logging the enzyme activites 
```{r}
killi_log<-killi  %>% mutate(across(c(CS_Activity, COX_Activity, LDH_Activity),
                ~log(.x),
                .names = "log_{.col}"))

str(killi_log)
```

Predictors: Acclimation temperautre, acclimation salinity, body condition. Body condtion is a continous predictor. 
```{r}

Manova_enzymes<- manova(cbind(log_COX_Activity,log_CS_Activity,log_LDH_Activity)~ Acclimation_Temperature*Acclimation_Salinity* Assay_Temperature+ Body_Condition, data=killi_log)

Anova(Manova_enzymes)

summary(Manova_enzymes)

```

Linear model for body condition with predictors temperature acclimation, salinity acclimation, and temperature assay. Wanted Fish ID but could not have it in the plot...
```{r}
body<- 
  lm(Body_Condition ~Acclimation_Temperature*Acclimation_Salinity , data=killi)

plot(body)

summary(body)

```


```
`

all effects plot 
```{r}
library(effects)

plot(allEffects(body))



(alleffects(body))

ggsave(
  filename = "bodyall.png",
  plot = allEffects(body),
  width = 8,  
  height = 5,
  dpi = 300
)
## see that temperature has a greater effect than the salinity, not expected as salinity should be an added factor to the body condition decrease 
plot(allEffects(body2))

```



Hematcrit 
```{r}

hem<- 
  lm(Hematocrit~Accliamtion_Temperature*Acclimation_Salinity*Assay_Temperature, data=killi)

plot(hem)


plot(allEffects(hem))

```




```