---
title: "MANCOVA"
author: "Hayden Simm"
date: "2026-04-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages
```{r}
library(tidyverse)
library(car)
library(heplots)
library(jmv)
library(effectsize)
library(patchwork)
```

Creating dataframe
```{r}
killi<-read_csv("../Data_Set/HS_MD_killifish.csv")

str(killi)
```

Changing predictors to factors 
```{r}
killi$Acclimation_Temperature<- factor (killi$Acclimation_Temperature)
killi$Acclimation_Salinity<- factor(killi$Acclimation_Salinity)
killi$Assay_Temperature<- factor(killi$Assay_Temperature)
str(killi)
```

Logging the enzyme activites 
```{r}
killi_log<-killi  %>% mutate(across(c(CS_Activity, COX_Activity, LDH_Activity),
                ~log(.x),
                .names = "log_{.col}"))

str(killi_log)
```

MANCOVA- because body condition would be a continuous predictor. 

Checking the assumption of no multicollinearity 
```{r}
cor(killi_log[, c("log_CS_Activity","log_COX_Activity","log_LDH_Activity")])
pairs(killi_log[, c("log_CS_Activity","log_COX_Activity","log_LDH_Activity")])
```

Checking assumption of regression slopes
```{r}
Group <- interaction(killi_log$Acclimation_Temperature,
                        killi_log$Acclimation_Salinity,
                        killi_log$Assay_Temperature)
                        
CS <- lm(log_CS_Activity ~ Group * Body_Condition, data = killi_log)
anova(CS)

COX <- lm(log_COX_Activity ~ Group * Body_Condition, data = killi_log)
anova(COX)

LDH <- lm(log_LDH_Activity ~ Group * Body_Condition, data = killi_log)
anova(LDH)
```
All pre-assumptions are met

Testing normality in responses and homogeneity of covariance
```{r}
jmv::mancova(data= killi_log, deps= c(log_COX_Activity,log_CS_Activity, log_LDH_Activity), factors= c(Acclimation_Temperature, Acclimation_Salinity, Assay_Temperature), multivar= list('pillai'),  covs = Body_Condition, boxM = TRUE,
  shapiro = TRUE, qqPlot = TRUE)
```
The qq-plot demonstrates the data follows normal distrubution even though shapiro wilks test says the data does not

```{r}
jmv::mancova(data= killi_log, deps= c(log_COX_Activity,log_CS_Activity, log_LDH_Activity), factors= c(Acclimation_Temperature, Acclimation_Salinity, Assay_Temperature), multivar= list('pillai'),  covs = Body_Condition, boxM = TRUE,
  shapiro = TRUE, qqPlot = TRUE)
```

Making the MANCOVA model
```{r}
mancova <- manova(
  cbind(log_COX_Activity, log_CS_Activity, log_LDH_Activity) ~
    Acclimation_Temperature * Acclimation_Salinity * Assay_Temperature +
    Body_Condition,
  data = killi_log
)
```

Testing interactions
```{r}
Anova(mancova, test.statistic = "Pillai")
```

Multivariate effect sizes
```{r}
eff <- eta_squared(mancova, partial = TRUE)
print(eff)
```
Effect size plot with thresholds
```{r}
panel1 <- ggplot(eff, aes(x = reorder(Parameter, Eta2_partial), y = Eta2_partial)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_hline(yintercept = 0.15, linetype = "dashed") +
  labs(
    x = "Predictor",
    y = "Partial eta²",
    title = "A") +
    theme_bw()
print(panel1)
```

Confidence interval plot
```{r}
panel2 <- ggplot(eff, aes(x = reorder(Parameter, Eta2_partial), y = Eta2_partial)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.2) +
  coord_flip() +
  labs(x = "Predictor", 
       y = "Partial eta²",
       title = "B") +
    theme_bw()
print(panel2)
```


```{r}
panel_plot <- panel1 / panel2

print(panel_plot)

ggsave(
  filename = "../Figures/mancova_effectsize.png",
  plot = panel_plot,
  width = 8,
  height = 8,
  dpi = 300
)
```

Univariate effect sizes
```{r}
aov_res <- summary.aov(mancova)
print(aov_res)

aov_res2<- eta_squared(aov(mancova), partial = TRUE)
print(aov_res2)
```

