---
title: "rda"
author: "Hayden Simm"
date: "2026-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages
```{r}
library(tidyverse)
library(vegan)
library(ggrepel)
```

Loading in dataframe and checking structure
```{r}
killi_rda <- read_csv("../Data_Set/HS_MD_killifish.csv")

str(killi_rda)
```
The predictors are all numeric, some need to be factors

Converting predictors to factors and Fish_ID to character
```{r}
killi_rda <- killi_rda %>%
  mutate(across(c(Assay_Temperature,
                  Acclimation_Temperature,
                  Acclimation_Salinity),
                as.factor))

killi_rda$Fish_ID <- as.character(killi_rda$Fish_ID)

str(killi_rda)
```


log transforming
```{r}
killi_rda <- killi_rda %>%
  mutate(across(c(CS_Activity, COX_Activity, LDH_Activity),
                ~log(.x),
                .names = "log_{.col}"))
```

Defining the response matrix and predictors
```{r}
response_matrix <- killi_rda %>%
  select(log_CS_Activity, log_COX_Activity, log_LDH_Activity)
```

Producing the RDA model
```{r}
rda_model <- rda(response_matrix ~ Assay_Temperature +
                                   Acclimation_Temperature +
                                   Acclimation_Salinity,
                 data = killi_rda)
```

Summary table
```{r}
print(
  summary(rda_model)
    )
```

Testing model
```{r}
print(
  anova(rda_model, permutations = how(blocks = killi_rda$Fish_ID))
    )
```
Overall model is significant so proceeding with interpretations by term. We used permutation-based RDA with Fish_ID as a blocking factor to account for repeated measures on the same fish across assay temperatures

RD1 explains almost all of the variance, so applying a significance test
```{r}
print(
  anova(rda_model, by = "axis",
      permutations = how(blocks = killi_rda$Fish_ID))
    )
```

Now doing a term wise test to see how individual predictors contribute to variation5
```{r}
print(
  anova(rda_model, by = "terms",
      permutations = how(blocks = killi_rda$Fish_ID))
    )
```

Extracting and joining scores for plotting
```{r}
sites <- scores(rda_model, display = "sites", scaling = 2)
species <- scores(rda_model, display = "species", scaling = 2)

sites <- as.data.frame(sites)
species <- as.data.frame(species)

species$label <- rownames(species)

sites$Assay_Temperature <- killi_rda$Assay_Temperature
sites$Acclimation_Temperature <- killi_rda$Acclimation_Temperature
sites$Acclimation_Salinity <- killi_rda$Acclimation_Salinity
```

Plot
```{r}
library(ggrepel)

ggplot(sites, aes(RDA1, RDA2)) +
  
  geom_point(aes(
    color = Assay_Temperature,
    fill = Acclimation_Salinity,
    shape = Acclimation_Temperature
  ),
  size = 2.5,
  stroke = 1.2,
  alpha = 0.9) +
  
  scale_color_manual(values = c("10" = "blue",
                                "24" = "red")) +
  
  scale_shape_manual(values = c("10" = 21,
                               "24" = 24)) +
  
  scale_fill_manual(values = c("28" = "grey40",
                              "11" = "white")) +
  
  # 🌿 RESPONSE VECTORS (arrows)
  geom_segment(data = species,
               aes(x = 0, y = 0,
                   xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "black") +
  
  geom_text_repel(data = species,
                  aes(x = RDA1, y = RDA2, label = label),
                  size = 4) +
  
  guides(
    fill = guide_legend(override.aes = list(shape = 21, colour = "black"))
  ) +
  
  theme_minimal() +
  coord_equal() +
  
  labs(
    color = "Assay Temp",
    shape = "Acclimation Temp",
    fill = "Acclimation Salinity"
  )
```

