---
title: "rda"
author: "Hayden Simm"
date: "2026-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages
```{r}
library(tidyverse)
library(vegan)
library(ggrepel)
library(patchwork)
```

Loading in dataframe and checking structure
```{r}
killi_rda <- read_csv("../Data_Set/HS_MD_killifish.csv")

str(killi_rda)
```
The predictors are all numeric, some need to be factors

Converting predictors to factors and Fish_ID to character
```{r}
killi_rda <- killi_rda %>%
  mutate(across(c(Assay_Temperature,
                  Acclimation_Temperature,
                  Acclimation_Salinity),
                as.factor))

killi_rda$Fish_ID <- as.character(killi_rda$Fish_ID)

str(killi_rda)
```


log transforming
```{r}
killi_rda <- killi_rda %>%
  mutate(across(c(CS_Activity, COX_Activity, LDH_Activity),
                ~log(.x),
                .names = "log_{.col}"))
```

Defining the response matrix and predictors
```{r}
response_matrix <- killi_rda %>%
  select(log_CS_Activity, log_COX_Activity, log_LDH_Activity)
```

Producing the RDA model
```{r}
rda_model <- rda(response_matrix ~ Assay_Temperature +
                                   Acclimation_Temperature +
                                   Acclimation_Salinity,
                 data = killi_rda)
```

RDA doesn't have the same diagnostic plots as a linear model but some plots to probe the model
```{r}

# Eigenvalues (RDA axes)
eig_df <- data.frame(
  Axis = seq_along(rda_model$CCA$eig),
  Eigenvalue = rda_model$CCA$eig
)

# Goodness-of-fit (sites)
gof_vals <- goodness(rda_model, display = "sites")
gof_df <- data.frame(
  Site = seq_along(gof_vals),
  GOF = as.numeric(gof_vals)
)

fitted_vals <- fitted(rda_model)
resid_vals <- residuals(rda_model)

fr_df <- data.frame(
  Fitted = fitted_vals[,1],
  Residuals = resid_vals[,1]
)

# ----------------------------
# Plot 1: Eigenvalues
# ----------------------------
p1 <- ggplot(eig_df, aes(x = Axis, y = Eigenvalue)) +
  geom_col() +
  labs(title = "A",
       x = "RDA axis",
       y = "Eigenvalue") +
  theme_bw()

# ----------------------------
# Plot 2: Goodness-of-fit
# ----------------------------
p2 <- ggplot(gof_df, aes(x = GOF)) +
  geom_histogram(bins = 30) +
  labs(title = "B",
       x = "GOF",
       y = "Count") +
  theme_bw()

# ----------------------------
# Plot 3: Fitted vs Residuals plot
# ----------------------------

p3 <- ggplot(fr_df, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "C",
       x = "Fitted values",
       y = "Residuals") +
  theme_bw()


# ----------------------------
# Combine into panel
# ----------------------------
panel_plot <- (p1 | p2) / (p3)

ggsave(
  filename = "../Figures/rda_diagnostics.png",
  plot = panel_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

Summary table
```{r}
print(
  summary(rda_model)
    )
```

Testing model
```{r}
print(
  anova(rda_model, permutations = how(blocks = killi_rda$Fish_ID))
    )
```
Overall model is significant so proceeding with interpretations by term. We used permutation-based RDA with Fish_ID as a blocking factor to account for repeated measures on the same fish across assay temperatures

RD1 appears to explain most of the variance, so applying a significance test and scree plot
```{r}
print(
  anova(rda_model, by = "axis",
      permutations = how(blocks = killi_rda$Fish_ID))
    )
```

Now doing a term wise test to see how individual predictors contribute to variation
```{r}
print(
  anova(rda_model, by = "terms",
      permutations = how(blocks = killi_rda$Fish_ID))
    )
```

Extracting and joining scores for analysis and plotting
```{r}
sites <- scores(rda_model, display = "sites", scaling = 2)
species <- scores(rda_model, display = "species", scaling = 2)
bp <- scores(rda_model, display = "bp", scaling = 2)

print(species)
print(bp)

sites <- as.data.frame(sites)
species <- as.data.frame(species)

species$label <- rownames(species)

sites$Assay_Temperature <- killi_rda$Assay_Temperature
sites$Acclimation_Temperature <- killi_rda$Acclimation_Temperature
sites$Acclimation_Salinity <- killi_rda$Acclimation_Salinity
```

RDA Biplot
```{r}
library(ggrepel)

rda_plot<-ggplot(sites, aes(RDA1, RDA2)) +
  
  geom_point(aes(
    color = Assay_Temperature,
    fill = Acclimation_Salinity,
    shape = Acclimation_Temperature
  ),
  size = 2.5,
  stroke = 1.2,
  alpha = 0.9) +
  
  scale_color_manual(values = c("10" = "blue",
                                "24" = "red")) +
  
  scale_shape_manual(values = c("10" = 21,
                               "24" = 24)) +
  
  scale_fill_manual(values = c("28" = "grey40",
                              "11" = "white")) +
  
  # RESPONSE VECTORS (arrows)
  geom_segment(data = species,
             aes(x = 0, y = 0,
                 xend = RDA1, yend = RDA2),
             linewidth = 1,
             arrow = arrow(length = unit(0.25, "cm")),
             color = "black") +
  
  geom_text_repel(data = species,
                  aes(x = RDA1, y = RDA2, label = label),
                  size = 5) +
  
  guides(
    fill = guide_legend(override.aes = list(shape = 21, colour = "black"))
  ) +
  
  theme_bw() +
  coord_equal() +
  
  labs(
    color = "Assay Temperature (°C)",
    shape = "Acclimation Temperature (°C)",
    fill = "Acclimation Salinity (ppt)"
  )
```

Saving RDA biplot
```{r}
ggsave(
  filename = "../Figures/rda_plot.png",
  plot = rda_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```
